---
title: "Using Optmatch on data in SAS, Stata, etc"
author: "Ben B. Hansen, Mark Fredrickson, Josh Buckner, Josh Errickson, and Peter Solenberger, with embedded Fortran code due to Dimitri P. Bertsekas and Paul Tseng"
date: "`r Sys.Date()`"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Using Optmatch with Foreign Software}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, prompt=TRUE)
```

# General comments

If a users preferred data analysis software is other than R, Optmatch
can still easily be used to perform the matching while all other data
analysis can be performed in the preferred software.

In general, the procedure will be

1. In your preferred software, export a data set containing a treatment indicator and all variables to match, exactMatch or caliper on.
    * (Optionally, if you wish to match using a propensity score, fit such a model and include the predicted propensity scores in the data set.)
1. Import the data into R and perform the matching.
1. Export the data from R, including the matches.
1. Import the data back into your preferred software.

The two best ways to import the data back-and-forth are either using
comma separated value files (.csv files) or using the R package `foreign`.

For .csv files, sample R code may be

```{r,eval=FALSE}
externaldata <- read.csv("externaldata.csv", header=TRUE)
externaldata$match <- fullmatch(..., data=externaldata)
write.csv(externaldata, file="externaldata.matched.csv")
```

Examples of the use of `foreign` are included within the
software-specific sections below.

# Using Optmatch with SAS

For this example, lets say we have some simple demographics. We will
treat gender as the treatment indicator, and wish to match on a
combination of a propensity score for gender (using both age and
height) and age.

```sas
data people;
  infile datalines dsd dlm=' ' missover;
  input gender age height;

datalines;
0 25 62
0 41 68
0 38 63
0 22 62
1 33 70
1 35 71
1 47 68
1 23 64
;
run;
```

Now we can fit a logistic model to predict gender using age and height.

```sas
proc logistic data = people;
  model gender (event='1') = age height;
  output out = preddata p=ppty;
run;
```

Finally, since we want to match only on the new `ppty` propensity
score and age, we can drop height.

```sas
proc data newpeople;
  set preddata;
  keep gender age ppty;
run;
```

With this setup, we can now either pass it to R via a .csv file, or
directly using the foreign package in R.

## Passing SAS data with .csv files

Save the file from SAS.

```sas
proc export data=newpeople;
  outfile="C:\Users\myuser\Desktop\sasout.csv";
run;
```

Inside R, we can load this data.

```{r,eval=FALSE}
sasdata <- read.csv("C:/Users/myuser/Desktop/sasout.csv", header=TRUE)
```

(Depending on your version of Windows, you may need to use

```{r,eval=FALSE}
sasdata <- read.csv("C:\\Users\\myuser\\Desktop\\sasout.csv", header=TRUE)
```

instead.)

If you have string variables (e.g. race as "White", "Hispanic", etc),
you may need to include the argument `stringsAsFactors=FALSE`.

Now, perform matching as desired, saving the final match to
`sasdata`. For example,

```{r,eval=FALSE}
library(optmatch)
f <- fullmatch(gender ~ age + ppty, data=sasdata)
sasdata$match <- f
```

Save this data back to .csv as follows.

```{r,eval=FALSE}
write.csv(sasdata, "C:/Users/myuser/Desktop/rout.sas.csv", row.names=FALSE)
```

The use of `row.names=FALSE` stops R from including the row names
(likely 1, 2, 3, etc) as the first column in the data. If you
re-arranged the data at any point, you may need to set that to `TRUE`,
but keep in mind to handle it properly in SAS, as the default will be
to treat it as a variable.

Now, returning to SAS, we can read the new rout.sas.csv file in. The only
catch is that we want to ensure that the match is read as a string by
using `$`, since it may have values like `1.1` and `1.10`,
representing two different matches, but which are identical if treated
as numeric.

```sas
data matchedpeople;
  infile "C:/Users/myuser/Desktop/rout.sas.csv" dsd firstobs=2;
  input gender age ppty match $;
run;
```

The argument `firstobs=2` skips the variable names; alternatively you
could pass `col.names=FALSE` to R's `write.csv`, but then the
rout.sas.csv file lacks any variable information, which may be useful
to have.

## Passing SAS data with R's foreign

As an alternative to using R's `write.csv`, you can use the R package
foreign to generate both the data and SAS code needed.

```{r,eval=FALSE}
library(foreign)
write.foreign(sasdata, 'C:/Users/myuser/Desktop/rout.sas.txt',
              'C:/Users/myuser/Desktop/rout.code.sas', package = 'SAS')
```

Opening the rout.code.sas file will give you the SAS code to read in
the data.

# Using Optmatch with Stata

For this example, we will start with the built-in auto data set in Stata.

```stata
sysuse auto.dta
```

We will treat `foreign`, whether a car is domestic or foreign, as the
treatment indicator. (Not to be confused with the R package foreign!)
We will estimate propensity scores using all other variables
(excluding `make` which is unique per row), and wish to match on the
estimated propensity score as well as `price` and `mpg`.

First, lets fit the logistic regression model.

```stata
logit foreign price mpg rep78 headroom trunk weight length turn displacement gear_ratio
predict ppty
```

Now, we can export a .csv file. Note that in addition to the treatment
indicator and variables to match on, we need to include a unique
identifier. In this case, we can use `make`. If no such identifier exists, you can use

```stata
gen case_id = _n
```

to generate ID's 1, 2, etc. These will be needed to merge the match information back in.

```stata
outsheet make foreign price mpg ppty using C:\Users\myuser\Desktop\stataout.csv, comma
```

Turning to R, this can be read in similar to the SAS example, using

```{r,eval=FALSE}
statadata <- read.csv("C:/Users/myuser/Desktop/stataout.csv", header=TRUE)
```

(Again, depending on your version of Windows, you may need to use

```{r,eval=FALSE}
statadata <- read.csv("C:\\Users\\myuser\\Desktop\\stataout.csv", header=TRUE)
```

instead.)

If you have string variables (e.g. race as "White", "Hispanic", etc),
you may need to include the argument `stringsAsFactors=FALSE`.

Now, perform matching as desired, saving the final match to
`statadata`. For example,

```{r,eval=FALSE}
library(optmatch)
f <- fullmatch(foreign ~ price + mpg + ppty, data=statadata, max.controls=3)
statadata$match <- f
```

We do not recommend using .csv files to transfer the data back to
Stata, though the `write.csv` file would be similar to that for
SAS. Instead, we recommend using foreign.

```{r,eval=FALSE}
write.dta(statadata, "C:/Users/myuser/Desktop/rout.stata.dta")
```

Back in Stata, you can merge this into the existing `auto.dta` by the
following commands.

```stata
sort make
merge 1:1 make using "C:/Users/myuser/Desktop/rout.stata.dta", force
```

The `force` option is necessary to overcome type
differences. Additional tweaks may be necessary here if you have
special variable types.
